---
title: "Class12"
author: "Aaron (PID A17544470)"
format: pdf
toc: true
---

# Section 1. Proportion of G/G in population

Download a CSV file from Ensemble.

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```


# Section 4. Population Scale Analysis

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. Hint: The read.table(),
summary() and boxplot() functions will likely be useful here. There is an example R script
online to be used ONLY if you are struggling in vein. Note that you can find the medium value
from saving the output of the boxplot() function to an R object and examining this object.
There is also the medium() and summary() function that you can use to check your
understanding.

There is 108 for A|A, 233 for A|G, and 121 for G|G

```{r}
url <- read.table("rs8067378_ENSG00000172057.6.txt")
```

```{r}
head(url)
summary(url)
nrow(url)
```

```{r}
table(url$geno)
```




> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)
```


```{r}
ggplot(url) + aes(geno, exp, fill = geno) +
  geom_boxplot(noth = T)
```





