---
title: "Comparative structural analysis"
author: "Aaron (PID A17544470)"
format: html
---

Here we run through a complete "pipeline" of a structure analysis that begins with a single sequence identifier and ends in a PCA analysis.

```{r}
library(bio3d)
```

## Step 1. retrieve a sequence for the protein we are interested in. We will take ADK "1ake_A"

```{r}
id <- "1ake_A"
aa <- get.seq(id)
aa
```

## Step 2. Run a BLAST search of the PDB for all related sequences to our input sequence "1ake_A"

```{r}
blast <- blast.pdb(aa)
```


```{r}
hits <- plot(blast)
```


## Step 3. Download related structures

These are our "top hits" i.e. all the structures in the PDB database related to our input sequence
```{r}
hits$pdb.id
```

```{r}
# Download related PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split = T, gzip = T)
```


## Step 4. Align and Super pose

```{r}
# Align related PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")

```

```{r}
pdbs
```


## Step 5. PCA

Let's use our old friend PCA to make sense of these confusing, complicated structure relationships.

```{r}
# Perform PCA
pc <- pca(pdbs)
plot(pc)
```

Let's make a trajectory or movie of the main conformational changeds captured by PC1 . We will use the `mktrj()` function for this...  

```{r}
mktrj(pc, file = "pca_result.pdb")
```



Back of the envelop comparison of the PDB size to UniProt

```{r}


uniprot <- 253206171
pdb <- 231029
pdb/uniprot * 100
```


Components -> polymer settings -> validation -> plDDT score
















